---
title: "Distributional Mediation Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dma)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Beyond the Mean: Why Distributional Mediation?

Traditional mediation analysis focuses on decomposing the **Average Treatment Effect (ATE)** into the **Natural Direct Effect (NDE)** and the **Natural Indirect Effect (NIE)**. While these provide a useful summary, they only describe the "center" of the causal response.

Causal mechanisms often shift the **shape**, **variance**, or **modality** of an outcome without necessarily shifting its mean. Energy Regression (`engression`) allows us to recover the entire counterfactual distribution $P(Y | do(A=d))$, giving us a much richer picture of the causal process.

## Simulation: The Multi-Modal Response

Consider a scenario where a treatment affects an outcome through a mediator that has a non-linear, squaring effect. Standard linear models or even non-linear mean models might capture the average shift but miss that the treatment introduces extreme values or multi-modality.

### Data Generating Process

- $A \in \{0, 1\}$ (Binary Treatment)
- $M | A \sim N(A, 0.5^2)$ (Mediator shifts its mean with treatment)
- $Y = M^2 + W + \epsilon$ (Outcome is a squared function of the mediator)
- $W \sim N(0, 1)$ (Confounder)

```{r dgp}
set.seed(42)
n <- 2000
W <- rnorm(n)
A <- rbinom(n, 1, 0.5)
M <- rnorm(n, A, 0.5)
Y <- M^2 + W + rnorm(n, 0, 0.5)
data <- data.frame(A, M, Y, W)

ggplot(data, aes(x = M, fill = factor(A))) +
  geom_density(alpha = 0.5) +
  labs(title = "Mediator Distribution by Treatment", fill = "Treatment (A)") +
  theme_minimal()
```

### Oracle Counterfactual Outcome Distributions

Since we know the DGP, we can compute the true potential outcome distributions under each treatment regime. This is what `plot_counterfactual_density()` aims to recover from data.

Under $do(A=a)$: $M \sim N(a, 0.25)$ and $Y = M^2 + W + \epsilon$, so the outcome distribution changes not just in location but in **shape** — the squared mediator introduces right-skew when $A=1$.

```{r oracle-outcomes}
set.seed(42)
n_oracle <- 10000
W_o <- rnorm(n_oracle)
eps_o <- rnorm(n_oracle, 0, 0.5)

# Y(0): do(A=0), M ~ N(0, 0.25)
M0 <- rnorm(n_oracle, 0, 0.5)
Y0 <- M0^2 + W_o + eps_o

# Y(1): do(A=1), M ~ N(1, 0.25)
M1 <- rnorm(n_oracle, 1, 0.5)
Y1 <- M1^2 + W_o + eps_o

# Y(1, M(0)): direct effect — treat but keep mediator at control level
Y10 <- M0^2 + W_o + eps_o  # same as Y0 (no direct effect of A on Y here)

# Y(0, M(1)): indirect effect — don't treat but shift mediator as if treated
Y01 <- M1^2 + W_o + eps_o  # same as Y1

oracle_df <- rbind(
  data.frame(Y = Y0, Regime = "Y(0, M(0)): Control"),
  data.frame(Y = Y1, Regime = "Y(1, M(1)): Treated"),
  data.frame(Y = Y01, Regime = "Y(0, M(1)): Indirect only")
)

ggplot(oracle_df, aes(x = Y, fill = Regime)) +
  geom_density(alpha = 0.4) +
  labs(title = "Oracle Counterfactual Outcome Distributions",
       subtitle = "Y = M² + W + eps: treatment shifts distribution shape, not just location",
       x = "Outcome (Y)", y = "Density") +
  theme_minimal()
```

Key observations:

- **Y(0, M(0))** (control): concentrated near zero, since $M \approx 0$ and $M^2 \approx 0$
- **Y(1, M(1))** (treated): shifted right with heavier right tail, since $M \approx 1$ and $M^2 \approx 1$
- **Y(0, M(1))** (indirect only): identical to treated here because $A$ only affects $Y$ through $M$ — the entire treatment effect is indirect
- The **NIE mean** is $E[Y(1,M(1))] - E[Y(1,M(0))] \approx$ `r round(mean(Y1) - mean(Y0), 2)`, but this single number hides the distributional shift

This is the target that `plot_counterfactual_density()` recovers from the trained `dma` model, without knowing the true DGP.

## Fitting dma

We fit the `dma` model. Note that we use a sufficient number of epochs to allow the neural networks to learn the squared relationship.

```{r fit, eval=FALSE}
res <- dma(
  data = data, trt = "A", outcome = "Y", mediators = "M", covar = "W",
  d0 = \(data, trt) 0, d1 = \(data, trt) 1,
  effect = "N",
  control = dma_control(
    crossfit_folds = 2L,
    num_epochs = 50L,
    batch_size = 128L,
    riesz_epochs = 10L
  )
)

print(res)
```

## Visualizing Results

### Coefficient Plot

The `plot()` method shows point estimates and confidence intervals for all estimated effects:

```{r coef-plot, eval=FALSE}
plot(res)
```

### Counterfactual Densities

Because `dma` saves the underlying generative models, we can visualize the full counterfactual densities of the outcome $Y$ under different treatment regimes using `plot_counterfactual_density()`.

Under the Natural Indirect Effect decomposition, we care about:

1. $P(Y(1, M(1)))$: The distribution if everyone was treated.
2. $P(Y(0, M(0)))$: The distribution if everyone was in the control group.

```{r distributional-plot, eval=FALSE}
plot_counterfactual_density(res, n_samples = 2)
```

### Analysis

In this example, $Y$ is a function of $M^2$:

- When $A=0$, $M \approx 0 \implies Y \approx 0$ (the distribution is centered near 0).
- When $A=1$, $M \approx 1 \implies Y \approx 1$ (the distribution shifts and its shape changes).

Traditional methods report the average shift (NIE), but `dma` lets you see how the entire probability mass of the outcome moves — invaluable for understanding treatment mechanisms in non-linear systems.

## Summary

The `dma` package provides:

1. **Valid Causal Estimates**: Doubly-robust estimates of mean effects (NDE, NIE, ATE) with observation weights propagated through all nuisance estimation.
2. **Distributional Insight**: Full counterfactual density visualization via saved `engression` models.
3. **Flexibility**: Robustness to non-linearities and non-Gaussian noise that would violate the assumptions of standard mediation software.
