---
title: "Causal Mediation with Energy Regression: dma vs crumble"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Mediation with Energy Regression: dma vs crumble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r parallel, eval=FALSE}
# Optional: enable parallel cross-fitting
library(future)
plan(multisession)
```

## Overview

`dma` implements the same causal mediation analysis framework as
[crumble](https://github.com/nt-williams/crumble) (Liu, Williams, Rudolph &
Díaz, 2024), but replaces `mlr3superlearner` with
[engression](https://cran.r-project.org/package=engression) (Shen &
Meinshausen, 2024) for all outcome regressions. Engression learns the full
conditional distribution $P(Y \mid X)$ via the energy score, rather than fitting
a point prediction $E[Y \mid X]$ via an ensemble of classical ML learners.

Both packages share the same estimation pipeline:

1. **$\theta$ outcome regressions** — nested pseudo-outcome cascade
   ($\theta^3 \to \theta^2 \to \theta^1$ for natural/organic effects)
2. **$\alpha$ density ratios** — cascaded Riesz representer estimation via neural
   networks with the Riesz loss
3. **EIF-based inference** — efficient influence function yields point estimates,
   standard errors, and confidence intervals via the `ife` package

The key difference:

| Component | crumble | dma |
|-----------|---------|-------------|
| $\theta$ (outcome regressions) | `mlr3superlearner` | `engression` |
| $\alpha$ (density ratios) | Custom Riesz NN | Same architecture, same loss |
| Cross-fitting | `origami` | `origami` |
| Inference | `ife` | `ife` |

## Data: Weight & Behavior

We use the `weight_behavior` dataset from the `mma` package, which is the
primary example in crumble's documentation.  The causal structure is:

- **Treatment** (`sports`): sports participation (binary: 1 or 2)
- **Outcome** (`bmi`): body mass index (continuous)
- **Mediators** (`exercises`, `overweigh`): exercise frequency and overweight status
- **Baseline confounders** (`age`, `sex`, `tvhours`): age, sex, TV hours

The causal question: *How much of the effect of sports participation on BMI
operates directly vs.\ through exercise and overweight status?*

```{r data}
library(mma)
data(weight_behavior)
weight_behavior <- na.omit(weight_behavior)
dim(weight_behavior)
head(weight_behavior[, c("sports", "bmi", "exercises", "overweigh", "age", "sex", "tvhours")])
```

## Shift Functions

Both packages use modified treatment policies to define interventions.
For binary/categorical treatments, we set everyone to the control or treatment
level:

```{r shifts}
d0 <- \(data, trt) factor(rep(1, nrow(data)), levels = c("1", "2"))
d1 <- \(data, trt) factor(rep(2, nrow(data)), levels = c("1", "2"))
```

## Natural Effects

Natural direct and indirect effects decompose the ATE into the pathway that
operates directly ($A \to Y$) and indirectly through the mediators ($A \to M \to Y$).

### crumble

```{r crumble-natural, cache=FALSE}
library(crumble)
set.seed(2345)

res_crumble_n <- crumble(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0,
  d1 = d1,
  effect = "N",
  learners = c("mean", "glm"),
  nn_module = sequential_module(),
  control = crumble_control(crossfit_folds = 1L, epochs = 20L)
)

print(res_crumble_n)
```

### dma

```{r dma-natural, cache=FALSE}
library(dma)
set.seed(2345)

res_dma_n <- dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0,
  d1 = d1,
  effect = "N",
  control = dma_control(
    crossfit_folds = 1L,
    num_epochs = 100L,
    riesz_epochs = 20L,
    noise_dim = 5L,
    hidden_dim = 100L,
    num_layer = 3L
  )
)

print(res_dma_n)
```

### Comparison

```{r compare-natural}
tidy_crumble_n <- data.frame(
  term = names(res_crumble_n$estimates),
  crumble_est = sapply(res_crumble_n$estimates, \(x) x@x),
  crumble_se = sapply(res_crumble_n$estimates, \(x) x@std_error)
)

tidy_enmed_n <- tidy(res_dma_n)

comparison_n <- merge(tidy_crumble_n, tidy_enmed_n[, c("term", "estimate", "std.error")],
                      by = "term", suffixes = c("_crumble", "_enmed"))
names(comparison_n)[4:5] <- c("enmed_est", "enmed_se")
print(comparison_n)
```

## Organic Effects

Organic effects decompose the ATE into:

- **ODE** (Organic Direct Effect): effect of $A$ on $Y$ *not* through $M$
- **OIE** (Organic Indirect Effect): effect of $A$ on $Y$ through $M$

Unlike natural effects, organic effects do not require a cross-world
counterfactual — they compare observed vs.\ natural mediator distributions.

### crumble

```{r crumble-organic, cache=FALSE}
set.seed(2345)

res_crumble_o <- crumble(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0,
  d1 = d1,
  effect = "O",
  learners = c("mean", "glm"),
  nn_module = sequential_module(),
  control = crumble_control(crossfit_folds = 1L, epochs = 20L)
)

print(res_crumble_o)
```

### dma

```{r dma-organic, cache=FALSE}
set.seed(2345)

res_dma_o <- dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0,
  d1 = d1,
  effect = "O",
  control = dma_control(
    crossfit_folds = 1L,
    num_epochs = 100L,
    riesz_epochs = 20L,
    noise_dim = 5L,
    hidden_dim = 100L,
    num_layer = 3L
  )
)

print(res_dma_o)
```

### Comparison

```{r compare-organic}
tidy_crumble_o <- data.frame(
  term = names(res_crumble_o$estimates),
  crumble_est = sapply(res_crumble_o$estimates, \(x) x@x),
  crumble_se = sapply(res_crumble_o$estimates, \(x) x@std_error)
)

tidy_enmed_o <- tidy(res_dma_o)

comparison_o <- merge(tidy_crumble_o, tidy_enmed_o[, c("term", "estimate", "std.error")],
                      by = "term", suffixes = c("_crumble", "_enmed"))
names(comparison_o)[4:5] <- c("enmed_est", "enmed_se")
print(comparison_o)
```

## Simulation Study

To validate statistical properties, we run a small Monte Carlo study with a
known DGP where true effects are calculable.

**DGP**: $W \sim N(0,1)$, $A \mid W \sim \text{Bernoulli}(\text{logit}^{-1}(0.5W))$,
$M = 0.5A + 0.3W + \varepsilon_M$, $Y = A + 0.8M + 0.4W + \varepsilon_Y$.

True natural effects: NDE = 1.0, NIE = $0.8 \times 0.5 = 0.4$, ATE = 1.4.

```{r simulation, cache=FALSE}
set.seed(123)
B <- 5
n <- 500

results <- data.frame()

for (b in 1:B) {
  W <- rnorm(n)
  A <- rbinom(n, 1, plogis(0.5 * W))
  M <- 0.5 * A + 0.3 * W + rnorm(n, sd = 0.5)
  Y <- A + 0.8 * M + 0.4 * W + rnorm(n, sd = 0.5)
  sim_data <- data.frame(W = W, A = A, M = M, Y = Y)

  ctrl <- dma_control(crossfit_folds = 2L, num_epochs = 100L,
    riesz_epochs = 30L, noise_dim = 3L, hidden_dim = 50L, num_layer = 2L)

  res <- dma(data = sim_data, trt = "A", outcome = "Y", mediators = "M",
    covar = "W", d0 = function(data, trt) 0, d1 = function(data, trt) 1,
    effect = "N", control = ctrl)

  td <- tidy(res)
  td$rep <- b
  results <- rbind(results, td)
}

# Summarize
true_vals <- c(direct = 1.0, indirect = 0.4, ate = 1.4)
summary_df <- do.call(rbind, lapply(names(true_vals), function(nm) {
  sub <- results[results$term == nm, ]
  data.frame(
    term = nm,
    true = true_vals[nm],
    mean_est = mean(sub$estimate),
    bias = mean(sub$estimate) - true_vals[nm],
    emp_se = sd(sub$estimate),
    mean_se = mean(sub$std.error),
    coverage = mean(sub$conf.low <= true_vals[nm] & sub$conf.high >= true_vals[nm])
  )
}))

rownames(summary_df) <- NULL
print(summary_df)
```

## Summary

Both `crumble` and `dma` implement the same semiparametric estimation
framework for causal mediation.  The key trade-off:

- **crumble** uses `mlr3superlearner`, which ensembles many classical ML
  learners (GLM, random forest, MARS, etc.) — flexible but requires many
  packages and tuning.
- **dma** uses `engression`, a single neural network that learns the
  full conditional distribution — simpler dependency structure, potentially
  better for complex conditional distributions, but requires torch.

For the Riesz representer ($\alpha$), both packages use the same approach:
a neural network trained with the Riesz loss $E[\alpha(X)^2 - 2 w \cdot f(\alpha, X')]$.

## Session Info

```{r session-info}
sessionInfo()
```
