---
title: "Effect Types: dma vs crumble"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect Types: dma vs crumble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

```{r parallel, eval=FALSE}
# Optional: enable parallel cross-fitting
library(future)
plan(multisession)
```

## Overview

dma supports four mediation effect decompositions from the general framework of Liu, Williams, Rudolph & Díaz (2024):

| Effect | Estimands | Requires `moc` |
|--------|-----------|:----------:|
| Natural (N) | NDE, NIE, ATE | No |
| Organic (O) | ODE, OIE | No |
| Randomized Interventional (RI) | RDE, RIE | Yes |
| Recanting Twins (RT) | 4-pathway decomposition | Yes |

This vignette demonstrates each effect type on the `weight_behavior` dataset from the `mma` package and compares results with [crumble](https://github.com/nt-williams/crumble), which uses the same estimation framework but with `mlr3superlearner` (ensemble of mean, GLM, MARS, and random forest) instead of engression for outcome regressions. Crumble results are taken from its [README](https://github.com/nt-williams/crumble) using `crossfit_folds = 1L, epochs = 20L` and `learners = c("mean", "glm", "earth", "ranger")`.

## Setup

```{r eval=FALSE}
library(dma)
library(mma)

data(weight_behavior)
weight_behavior <- na.omit(weight_behavior)

set.seed(2345)
```

The shift functions define the modified treatment policies — here, static interventions setting everyone to `sports = 1` (control) or `sports = 2` (treated):

```{r eval=FALSE}
d0 <- \(data, trt) factor(rep(1, nrow(data)), levels = c("1", "2"))
d1 <- \(data, trt) factor(rep(2, nrow(data)), levels = c("1", "2"))
```

## Natural Effects

Natural direct and indirect effects (Pearl, 2001). The NDE captures the effect of shifting treatment while holding the mediator at its natural value under the control regime; the NIE captures the remaining indirect pathway.

No mediator-outcome confounders (`moc`) are needed — the cross-world counterfactual is handled by the nested pseudo-outcome cascade.

```{r eval=FALSE}
dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0, d1 = d1,
  effect = "N",
  control = dma_control(crossfit_folds = 1L, num_epochs = 100L, riesz_epochs = 20L)
)
#> ✔ Fitting outcome regressions... 1/1 folds [40.8s]
#> ✔ Computing alpha n density ratios... 1/1 folds [40.5s]
#>
#> ══ Results `dma()` ═════════════════════════════════════════
#>
#> ── Direct Effect
#>       Estimate: -0.043
#>     Std. error: 0.18
#> 95% Conf. int.: -0.396, 0.31
#>
#> ── Indirect Effect
#>       Estimate: 0.835
#>     Std. error: 0.232
#> 95% Conf. int.: 0.379, 1.29
#>
#> ── Average Treatment Effect
#>       Estimate: 0.792
#>     Std. error: 0.305
#> 95% Conf. int.: 0.195, 1.389
```

**Comparison with crumble:**

| Estimand | dma | crumble | Difference |
|----------|:-----------:|:-------:|:----------:|
| ATE      | 0.792 (0.305) | 1.029 (0.280) | -0.237 |
| NDE      | -0.043 (0.180) | 0.017 (0.172) | -0.060 |
| NIE      | 0.835 (0.232) | 1.012 (0.218) | -0.177 |

Both packages agree on the qualitative story: a small, insignificant direct effect and a large, significant indirect effect. Point estimates differ modestly — expected since engression (distributional regression) and mlr3superlearner (ensemble of classical learners) are fundamentally different learning strategies, and both runs use only 1 cross-fitting fold.

## Organic Effects

Organic direct and indirect effects (Lok, 2015). Unlike natural effects, organic effects use the *observed* mediator distribution rather than a counterfactual one, avoiding cross-world assumptions.

```{r eval=FALSE}
dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = d0, d1 = d1,
  effect = "O",
  control = dma_control(crossfit_folds = 1L, num_epochs = 100L, riesz_epochs = 20L)
)
#> ✔ Fitting outcome regressions... 1/1 folds [41s]
#> ✔ Computing alpha n density ratios... 1/1 folds [40.1s]
#>
#> ══ Results `dma()` ═════════════════════════════════════════
#>
#> ── Organic Direct Effect
#>       Estimate: -0.085
#>     Std. error: 0.177
#> 95% Conf. int.: -0.431, 0.262
#>
#> ── Organic Indirect Effect
#>       Estimate: 0.898
#>     Std. error: 0.233
#> 95% Conf. int.: 0.442, 1.354
```

**Comparison with crumble:**

| Estimand | dma | crumble | Difference |
|----------|:-----------:|:-------:|:----------:|
| ODE      | -0.085 (0.177) | 0.011 (0.177) | -0.096 |
| OIE      | 0.898 (0.233) | 1.028 (0.223) | -0.130 |

Again, qualitative agreement — insignificant direct effect, significant indirect effect — with comparable standard errors.

## Randomized Interventional Effects

Randomized interventional direct and indirect effects (Vansteelandt & Daniel, 2017). These effects are identified even in the presence of mediator-outcome confounders (`moc`) affected by treatment, at the cost of a weaker decomposition that uses a *randomized* mediator draw from the population.

The `moc` argument specifies intermediate confounders — here, `snack`. Estimation requires Z' permutation via Rsymphony.

```{r eval=FALSE}
dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  moc = "snack",
  d0 = d0, d1 = d1,
  effect = "RI",
  control = dma_control(crossfit_folds = 1L, num_epochs = 100L, riesz_epochs = 20L)
)
#> ✔ Permuting Z-prime variables... 1/1 tasks [2.1s]
#> ✔ Fitting outcome regressions... 1/1 folds [1m 0.7s]
#> ✔ Computing alpha r density ratios... 1/1 folds [1m 16.3s]
#>
#> ══ Results `dma()` ═════════════════════════════════════════
#>
#> ── Randomized Direct Effect
#>       Estimate: 0.014
#>     Std. error: 0.182
#> 95% Conf. int.: -0.343, 0.371
#>
#> ── Randomized Indirect Effect
#>       Estimate: 0.951
#>     Std. error: 0.245
#> 95% Conf. int.: 0.47, 1.431
```

**Comparison with crumble:**

| Estimand | dma | crumble | Difference |
|----------|:-----------:|:-------:|:----------:|
| RDE      | 0.014 (0.182) | 0.016 (0.177) | -0.002 |
| RIE      | 0.951 (0.245) | 1.030 (0.230) | -0.079 |

The closest agreement of any effect type — direct effects are nearly identical, and indirect effects are within one standard error.

## Recanting Twins

Recanting twins pathway decomposition (Vo et al., 2024). This is the finest decomposition, separating the total effect into four causal pathways plus an intermediate confounding term:

- **A → Y**: direct effect bypassing mediators and confounders
- **A → Z → Y**: effect through mediator-outcome confounders only
- **A → Z → M → Y**: effect through confounders then mediators
- **A → M → Y**: effect through mediators only
- **Intermediate confounding**: residual from confounders

```{r eval=FALSE}
dma(
  data = weight_behavior,
  trt = "sports",
  outcome = "bmi",
  covar = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  moc = "snack",
  d0 = d0, d1 = d1,
  effect = "RT",
  control = dma_control(crossfit_folds = 1L, num_epochs = 100L, riesz_epochs = 20L)
)
#> ✔ Permuting Z-prime variables... 1/1 tasks [2.1s]
#> ✔ Fitting outcome regressions... 1/1 folds [1m 46.8s]
#> ✔ Computing alpha n density ratios... 1/1 folds [1m 17.1s]
#> ✔ Computing alpha r density ratios... 1/1 folds [1m 17.5s]
#>
#> ══ Results `dma()` ═════════════════════════════════════════
#>
#> ── Path: A -> Y
#>       Estimate: 0.241
#>     Std. error: 0.191
#> 95% Conf. int.: -0.133, 0.615
#>
#> ── Path: A -> Z -> Y
#>       Estimate: 0.074
#>     Std. error: 0.02
#> 95% Conf. int.: 0.035, 0.113
#>
#> ── Path: A -> Z -> M -> Y
#>       Estimate: -0.019
#>     Std. error: 0.019
#> 95% Conf. int.: -0.055, 0.018
#>
#> ── Path: A -> M -> Y
#>       Estimate: 0.866
#>     Std. error: 0.232
#> 95% Conf. int.: 0.412, 1.32
#>
#> ── Intermediate Confounding
#>       Estimate: -0.208
#>     Std. error: 0.052
#> 95% Conf. int.: -0.311, -0.106
```

**Comparison with crumble:**

| Pathway | dma | crumble | Difference |
|---------|:-----------:|:-------:|:----------:|
| A → Y                | 0.241 (0.191) | 0.037 (0.184) | 0.204 |
| A → Z → Y            | 0.074 (0.020) | -0.020 (0.024) | 0.094 |
| A → Z → M → Y        | -0.019 (0.019) | -0.001 (0.010) | -0.018 |
| A → M → Y            | 0.866 (0.232) | 1.051 (0.216) | -0.185 |
| Intermediate conf.   | -0.208 (0.052) | -0.013 (0.026) | -0.195 |

The dominant pathway (A → M → Y) is consistent across both packages. The finer decomposition shows more variation between learners, which is expected — the RT decomposition involves the most nuisance parameters (both natural and randomized cascades), so differences in the outcome regression learner compound through more levels. Both packages agree that the A → M → Y path is by far the largest and most significant component.

## Summary

```{r echo=FALSE}
tab <- data.frame(
  Effect = c("Natural", "", "", "Organic", "", "Rand. Interventional", "",
             "Recanting Twins", "", "", "", "", ""),
  Estimand = c("ATE", "NDE", "NIE", "ODE", "OIE", "RDE", "RIE",
               "A->Y", "A->Z->Y", "A->Z->M->Y", "A->M->Y", "Int. Conf.", "ATE"),
  dma = c("0.792", "-0.043", "0.835", "-0.085", "0.898", "0.014", "0.951",
                   "0.241", "0.074", "-0.019", "0.866", "-0.208", "0.954"),
  crumble = c("1.029", "0.017", "1.012", "0.011", "1.028", "0.016", "1.030",
              "0.037", "-0.020", "-0.001", "1.051", "-0.013", "1.054")
)
knitr::kable(tab, col.names = c("Effect", "Estimand", "dma", "crumble"),
             align = c("l", "l", "r", "r"),
             caption = "Point estimates on weight_behavior data (crossfit_folds = 1)")
```

Both packages target the same estimands using the same semiparametric framework, but differ in how nuisance parameters are estimated:

- **crumble** uses `mlr3superlearner` — an ensemble of classical ML learners (mean, GLM, MARS, random forest) — for outcome regressions
- **dma** uses `engression` — a single neural network trained with the energy score to learn the full conditional distribution

Qualitative conclusions are identical across all four effect types: a large, significant indirect effect operating through the mediators (`exercises`, `overweigh`), with a small and generally insignificant direct effect. Point estimate differences reflect the different learners and the inherent variability from using a single cross-fitting fold.

## References

- Liu, R., Williams, N., Rudolph, K., & Díaz, I. (2024). General Mediation Analysis. arXiv:2408.14620.
- Lok, J. J. (2015). Defining and estimating causal direct and indirect effects when setting the mediator to specific values is not feasible. *Statistics in Medicine*.
- Pearl, J. (2001). Direct and indirect effects. *Proceedings of UAI*.
- Shen, X. & Meinshausen, N. (2024). Engression: Extrapolation through the Lens of Distributional Regression. *JMLR*.
- Vansteelandt, S. & Daniel, R. M. (2017). Interventional effects for mediation analysis with multiple mediators. *Epidemiology*.
- Vo, T.-T. et al. (2024). An axiomatic approach to the decomposition of treatment effects using mediators. arXiv:2401.07574.
